---
title: "COVID19_Project"
author: "Julian, Nabih, Nathan, and Yehya"
date: 02-10-2020
output:
  github_document:
    toc: true
---

```{r setup}
library(tidyverse)
```

*Background*:


# Getting the Data
<!------------------------------------------------------>
##The GDP data can be substituted for GapMinder
```{r load-covid-data}
url_covid <- "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv"

## Set the filenames of the data to download
filename_covid <- "./data/covid_countries.csv"
filename_gdp <- "./data/gdp_countries.csv"

## Download the data locally
curl::curl_download(
        url_covid,
        destfile = filename_covid
)

## Loads the downloaded csv
df_covid_temp <- read_csv(filename_covid) %>% 
  select(iso_code, continent, date, total_cases, total_deaths, total_cases_per_million, total_deaths_per_million, population)

df_gdp_temp <- read_csv(filename_gdp) %>% 
  select("Country Name", "2019","Country Code")
##Rename columns:
df_gdp <- df_gdp_temp %>% 
  rename(
    iso_code = "Country Code",
    country = "Country Name",
    gdp = "2019"
    )
##Rename columns:
df_covid <- df_covid_temp %>% 
  rename(
    "cases" = total_cases,
    "deaths" = total_deaths,
    "cases_per1M" = total_cases_per_million,
    "deaths_per1M" = total_deaths_per_million
    )

## Summaries to check info is okay
summary(df_covid)

summary(df_gdp)
```

# Join the Data
<!-- -------------------------------------------------- -->
##Hopefully well be able to join the data using the ISO code, if not we can use the country name and see what problems that may bring
```{r join-covid-data}

df_data_raw <-df_gdp %>%
  ## Join both databases we have
            left_join(
              df_covid,
              by = "iso_code",
              na.rm = TRUE
            ) %>% 
  ## Whenever teh value of GDP is NA, we drop the country because we cant compare it to others. If teh value of a case/death count is NA, we set it to 0
              drop_na(gdp) %>% 
                mutate(cases = ifelse(is.na(cases), 0, cases)) %>% 
                  mutate(deaths = ifelse(is.na(deaths), 0, deaths)) %>% 
                    mutate(cases_per1M = ifelse(is.na(cases_per1M), 0, cases_per1M)) %>% 
                      mutate(deaths_per1M = ifelse(is.na(deaths_per1M), 0, deaths_per1M))

df_data_raw
```


# Work on the Data
<!-- -------------------------------------------------- -->

We are going to calculate GDP per capita so that our measurements to compare the countries are standardized, and we can get more "honest" measurements and comparisons.
```{r normalize-gdp-data}

df_data_norm <- df_data_raw %>% 
                  mutate(gdpPerCap = (gdp/population)) %>% 
                    select(continent, country, gdpPerCap, date, cases_per1M, deaths_per1M)

df_data_norm
```


```{r SA-plot}
df_data_norm %>%
  filter(continent == "South America") %>% 
  ggplot(
    aes(date, cases_per1M, color = fct_reorder2(country, date, cases_per1M))
  ) +
  geom_line() +
  ##scale_y_log10(labels = scales::label_number_si()) +
  scale_color_discrete(name = "Country") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Cases (per 1,000,000 persons)"
  )
```
```{r Asia-plot}
df_data_norm %>%
  filter(continent == "Asia") %>% 
  ggplot(
    aes(date, cases_per1M, color = fct_reorder2(country, date, cases_per1M))
  ) +
  geom_line() +
  ##scale_y_log10(labels = scales::label_number_si()) +
  scale_color_discrete(name = "Country") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Cases (per 1,000,000 persons)"
  )
```

```{r Africa-plot}
df_data_norm %>%
  filter(continent == "Africa") %>% 
  ggplot(
    aes(date, cases_per1M, color = fct_reorder2(country, date, cases_per1M))
  ) +
  geom_line() +
  ##scale_y_log10(labels = scales::label_number_si()) +
  scale_color_discrete(name = "Country") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Cases (per 1,000,000 persons)"
  )
```

```{r Europe-plot}
df_data_norm %>%
  filter(continent == "Europe") %>% 
  ggplot(
    aes(date, cases_per1M, color = fct_reorder2(country, date, cases_per1M))
  ) +
  geom_line() +
  ##scale_y_log10(labels = scales::label_number_si()) +
  scale_color_discrete(name = "Country") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Cases (per 1,000,000 persons)"
  )
```

```{r NA-plot}
df_data_norm %>%
  filter(continent == "North America") %>% 
  ggplot(
    aes(date, cases_per1M, color = fct_reorder2(country, date, cases_per1M))
  ) +
  geom_line() +
  ##scale_y_log10(labels = scales::label_number_si()) +
  scale_color_discrete(name = "Country") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Cases (per 1,000,000 persons)"
  )
```
# Conclusion
<!-- -------------------------------------------------- -->