---
title: "COVID19_Report"
author: "Julian, Nabih, Nathan, and Yehya"
date: 20-10-2020
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
library(tidyverse)
```

<!-- -------------------------------------------------- -->
# *Background*

COVID-19 is a newly discovered (in 2019) virus in humans that mainly attacks the respiratory system. It has been thought to have mutated from the severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2) virus. Some statistics about COVID-19 (according to the CDC): around 14%-19% of people with COVID are hospitalized and 3%-5% of people need to go to the ICU (Intensive Care Unit). The main reason for its quick spread to a pandemic is the fact that it has an incubation period of 2-14 days and many people do not develop symptoms during that period. The risks from contracting COVID depend on age group and pre-existing conditions. Older age groups and people with pre-existing conditions are more likely to experience complications with COVID.

According to Our World in Data, countries with higher GDP tend to spend more money on PPP and healthcare. This then begs the question, what is the relationship between contraction/mortality rates of COVID-19 and the GDP of a country. We have developed a hypothesis and aim to discover the truth in this study.

<!-- -------------------------------------------------- -->
# *Question*

We set out to answer the question of whether or not the per capita GDP of nations around the world had any relationship with the total number of reported cases (per million persons) of the respective nations. We were wondering if the proportional wealth of each nation could have an affect on the number of cases each nation has. We hypothesized that if a country had more wealth per person that they would most likely have access to better education and resources to better prepare for a pandemic and thus would have fewer cases.

<!-- -------------------------------------------------- -->
# *Data*

For our project, we have two main sources of data: our COVID-19 dataset, and our GDP dataset. The COVID-19 dataset we got from [this GitHub repo](https://github.com/owid/covid-19-data/tree/master/public/data) create by [Our World in Data](https://ourworldindata.org/)  a group of researcher who seek to research and develop data to make progress against the world’s largest problems. Data produced by this group has been used for teaching at top universities like Harvard and Oxford, and for research in over 500 different papers and articles, from sources ranging from The New York Times to the Science Journal. This dataset is updated daily to show how our COVID-19 cases and deaths are developing worldwide, so each time our code is run the data will change a little because of the new datapoints added. Our other dataset is coming from [The World Bank](https://data.worldbank.org/indicator/NY.GDP.MKTP.CD) and gives us the most recent GDP values for different countries around the world, which as of now is the 2019 value.

We are joining these two datasets by using each country’s ISO code. This is a 3 letter unique code that represents each country. Besides that, we also have what continent each country belongs to (which is given by both datasets). Our COVID-19 dataset is also giving us Total Cases, Total Deaths, Cases per million persons, Deaths per million persons, and population of each country, with the first 4 of those values also coming with the date when they were reported. Our GDP dataset is giving us the GDP for each country, with the other columns being used to line it up to its COVID-19 counterpart. One important caveat of the COVID-19 dataset: values that have not been reported show up as N/A instead of 0, so we replaced all these values to be able to have an understandable graph for the same time spans for all our countries.



### Getting the Data and tidying the data
<!------------------------------------------------------>
```{r load-covid-data, message=FALSE, warning=FALSE}
url_covid <- "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/owid-covid-data.csv"

## Set the filenames of the data to download
filename_covid <- "./data/covid_countries.csv"
filename_gdp <- "./data/gdp_countries.csv"

## Download the data locally
curl::curl_download(
        url_covid,
        destfile = filename_covid
)

## Loads the downloaded csv
df_covid_temp <- read_csv(filename_covid) %>% 
  select(iso_code, continent, date, total_cases, total_deaths, total_cases_per_million, total_deaths_per_million, population)

df_gdp_temp <- read_csv(filename_gdp) %>% 
  select("Country Name", "2019","Country Code")
##Rename columns:
df_gdp <- df_gdp_temp %>% 
  rename(
    iso_code = "Country Code",
    country = "Country Name",
    gdp = "2019"
    )
##Rename columns:
df_covid <- df_covid_temp %>% 
  rename(
    "cases" = total_cases,
    "deaths" = total_deaths,
    "cases_per1M" = total_cases_per_million,
    "deaths_per1M" = total_deaths_per_million
    )
```

### Join the Data
<!-- -------------------------------------------------- -->
We are going to calculate GDP per capita so that our measurements to compare the countries are standardized, and we can get more "honest" measurements and comparisons.
```{r join-covid-data}

df_data_raw <-df_gdp %>%
            left_join(
              df_covid,
              by = "iso_code",
              na.rm = TRUE
            ) %>% 
              drop_na(gdp) %>% 
                drop_na(continent) %>% 
                  replace_na(list("cases" = 0, "deaths" = 0, "cases_per1M" = 0,  "deaths_per1M" = 0))

df_data_norm <- df_data_raw %>% 
                  mutate(gdpPerCap = (gdp/population), na.rm = TRUE) %>% 
                    select(continent, country, gdpPerCap, date, cases_per1M, deaths_per1M)
```

<!-- -------------------------------------------------- -->
# *Graphs and Observations*

```{r}
## Plot with low gdp countries compared to high gdp
df_data_norm %>% 
  filter(date == "2020-10-21") %>% 
  mutate(gdpPerCap = gdpPerCap/1000) %>% 
  arrange(gdpPerCap) %>% 
  mutate(country = fct_reorder(country, desc(gdpPerCap))) %>% 
  mutate( bin=cut_width(gdpPerCap, width=10, boundary=0) ) %>%
  ggplot( aes(x=bin, y=cases_per1M, group=bin)) +
  geom_boxplot(fill="#69b3a2") +
  theme_minimal() +
  coord_flip() +
  labs(
    x = "GDP per capita grouped by bins of 10,000",
    y = "Covid Cases per 1M people",
    title = "Reported Covid cases versus GDP per capita for each country"
  )
```

The above figure shows the number of reported COVID cases versus the gdp per capita of the country.

- From this graph we notice several interesting things:

- Case counts per 1m people appear to increase as GDP increases, this is contradictory to our hypothesis

- Lowest GDP bracket countries have the lowest reported cases

- The 60-70,000 GDP bracket looks like an outlier


```{r byCountry_nonscaled}
df_data_norm %>%
  filter(date == max(date)) %>% 
  ggplot(
    aes(gdpPerCap, cases_per1M)
  ) +
  geom_point() +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 1, raw=TRUE),colour="red") +
  facet_wrap(~ continent) +
  theme_minimal() +
  labs(
    x = "GDP Per Capita",
    y = "Cases (per 1,000,000 persons)"
  )
```

The above plot shows the most recent total reported cases per 1 million persons against the per capita GDP’s of the countries of the world. We separated the countries into their respective continents and placed a simple linear trend line to see if any possible relationships exist. Upon initial glance, regions such as Asia, North America, and Europe appear to have positive relationships even if they are more slight. But for the other regions it is much more difficult to pick out if the data points to an actual trend or not due to different ranges of values. So we decided to try and reformat the plot to be more clear.

``` {r byCountry_scaled}
df_data_norm %>%
  filter(date == max(date)) %>% 
  ggplot(
    aes(gdpPerCap, cases_per1M)
  ) +
  geom_point() +
  stat_smooth(method="lm", se=TRUE, fill=NA,
                formula=y ~ poly(x, 1, raw=TRUE),colour="red") +
  facet_wrap(~ continent, scale = "free") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45)
  ) +
  labs(
    x = "GDP Per Capita",
    y = "Cases (per 1,000,000 persons)"
  )
```

By allowing each cell’s axis-scaling to be different from one another, we can now more clearly see relationships within each continent. It appears that there are positive relationships in each region, meaning that there appears to be a relationship between higher GDP per capita and higher total reported cases. It is important to note that a fair amount of variability exists in many of the regions and others like Oceania have very few data points, so while there does appear to be positive trends, we might require more data to find a more decisive relationship.

<!-- -------------------------------------------------- -->
# *Conclusion*

In general, when looking at our data, we can see that our hypothesis was completely wrong. We believed that lower GDP would have a correlation to higher cases/deaths due to COVID-19. Looking at our data we got to a very different conclusion. When looking at the whole world, there appears to be a trend that higher GDP leads to an average of more cases, but then when broken down into continents we can see this conclusion break completely. Besides that, when separated into continents we can see how variable and different each continent was affected by the virus. First of all we see how continents like Africa and South America have extremely lower GDP than others like Asia, Europe, and North America, and we also see how this doesn't mean they have similar COVID-19 cases. South America has much more cases, both in average and in total, even though its GDP distribution is very similar to Africa. 

Our belief is that this variability by country is due to three factors: first, how well the continents and its countries reacted to the virus. Second, how early the continent was hit (we can see that Asia has more cases than North America, which might be because it was ground zero). And third, how much travel to that continent occurs on a normal basis. We believe that there is more transit between European countries than there is between African countries, which might explain why Africa has such a small number of cases.

It is also important to note that continents like Oceania and South America (especially the first) are hard to compare against others because of the small amount of data points we have for them, meaning that they can become outliers much easier, and can lead to harder conclusions.

In conclusion, that while we believe that there is a slight relationship between GDP and the amount of cases from COVID-19 a country has, the small relationship there might be is outweighed by other factors (such as how the country reacted to the virus, or how much travel there normally is to and from that country) making it close to insignificant in determining the effect of COVID-19 on a country.

<!-- -------------------------------------------------- -->
# *Questions for the future*

The above conclusions provide an answer to our hypothesis but they also spur us towards other questions. Some of the questions and extensions that we are interested in include:

-  Examining the relationship between gdp per capita at a state and county level in the US. We are curious if the trends we saw at a global level are also present at other populations sizes. Are these trends consistent between different countries?

- What unites, if anything, the outliers in our gdp / covid graph. For example maybe smaller or larger countries follow very different trends.

- Are there better metrics to look at how well different nations dealt with COVID? For example, counting cases likely exhibits lots of reporting bias etc. Maybe looking at inflation or gold prices etc within a country could explain which countries were more stable.

- We were also interested in the death/case ratio (look at our source if interested)

```{r include=FALSE,echo=FALSE}
## Plot with low gdp countries compared to high gdp
df_data_norm %>% 
  filter(date == "2020-10-21") %>% 
  mutate(gdpPerCap = gdpPerCap/1000) %>% 
  arrange(gdpPerCap) %>% 
  mutate(country = fct_reorder(country, desc(gdpPerCap))) %>% 
  mutate( bin=cut_width(gdpPerCap, width=10, boundary=0) ) %>%
  ggplot( aes(x=bin, y=deaths_per1M/cases_per1M, group=bin)) +
  geom_boxplot(fill="#69b3a2") +
  theme_minimal() +
  coord_flip() +
  labs(
    x = "GDP per capita grouped by bins of 10,000",
    y = "Deaths/Case ratio",
    title = "Reported Covid Deaths/Cases vs GDP/capita for each country"
  )
```
```{r include=FALSE, echo=FALSE}
# - This shows the number of reported deaths/cases ratio across different GDP brackets. When compared to the case incidence rate there are some interesting behaviors.
# 
# - Overall the shape of the death/case curve is very different than just the case curve
# 
# - It appears that the 60-70,000 gdp bracket has the lowest average death/case rate even though they had one of the higher case rates
# 
# - The lower gdp brackets had low case counts but higher deaths/case
# 
# - This leads to a larger design question of whether deaths per case may be a better metric to evaluate how countries were affected by the pandemic rather than just reported cases
```

